### 2023-05-26 submission/manuscript.tex
Add abstract / edits

\({\dagger}\) To whom corresponsdence should be addressed:

@@

@@

Microbiome studies that use amplicon sequencing to characterize multiple
samples use PCR to amplify 16S rRNA gene fragments using primers with
distinct barcodes or index sequences {[}(1); \textbf{XXXXX}{]}. These
barcodes allow researchers to pool PCR products and then deconvolute the
resulting sequence data based on the barcode sequences. Despite their
best efforts to generate equimolar pools of PCR products, it is common
to observe more than 10-fold variation in the number of sequences per
sample {[}\textbf{XXXXX}{]}. Researchers desire strategies to minimize
uneven sequencing depth and thus need methods to control for this
uneveness in their analyses. Of course, uneven sampling is not unique to
microbiome research and is a challenge faced by all community ecologists
{[}\textbf{XXXXX}{]}. Common approaches to controlling uneven sampling
efforts include use of proportional abundance (i.e., relative
abundance), scaling of counts, parameter estimation, and rarefaction
{[}\textbf{XXXXX}{]}.

In 2014 Paul McMurdie and Susan Holmes published ``Waste not, want not:
why rarefying microbiome data is inadmissible'' (WNWN) in \emph{PLOS
Computational Biology} (2). Their paper has had a significant impact on
the approaches that microbiome researchers use to analyze 16S rRNA gene
sequence data. According to Google Scholar, this paper has been cited
more than 2,400 times as of June 2023. There has been a rebuttal of WNWN
that actually showed how rarefaction could be beneficial (3).
Anecdotely, I have received correspondence from researchers over the
past 10 years asking how to address critiques from reviewers who
criticize my correspondents' analysis for using rarefaction (e.g., see
\href{https://twitter.com/inanna_nalytica/status/1264299305067786242}{this
Twitter thread}). I have also received similar comments from reviewers
regarding my own work. Most recently, I received the critique for a
preprint that I posted in 2020 that analyzed the practice of removing
rare taxa from microbiome analyses {[}see comments linked to
\textbf{XXXXX};
\url{https://web.archive.org/web/20230308191505/http://dalmug.org/filter-review/}{]}.
In the process of responding to these reviewers' comments and preparing
a manuscript investigating rarefaction and other approaches to control
for uneven sequencing effort, I grew to appreaciate that there is
signifcant confusion in the field over what is meant by ``rarefying''
and ``rarefaction''.

@@

\begin{quote}
``Instead, microbiome analysis workflows often begin with an ad hoc
library size normalization by random subsampling without replacement, or
so-called rarefying (4--6). There is confusion in the literature
regarding terminology, and sometimes this normalization approach is
conflated with a non-parametric resampling technique --- called
rarefaction (7), or individual-based taxon re-sampling curves (8) ---
that can be justified for coverage analysis or species richness
estimation in some settings (8), though in other settings it can perform
worse than parametric methods (9). Here we emphasize the distinction
between taxon re-sampling curves and normalization by strictly adhering
to the terms rarefying or rarefied counts when referring to the
normalization procedure, respecting the original definition for
rarefaction. Rarefying is most often defined by the following steps (5).

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Select a minimum library size, N\textsubscript{L,m}. This has also
  been called the rarefaction level (4), though we will not use the term
  here.
\item
  Discard libraries (microbiome samples) that have fewer reads than
  N\textsubscript{L,m}.
\item
  Subsample the remaining libraries without replacement such that they
  all have size N\textsubscript{L,m}.
\end{enumerate}

@@

It was unfortunate that McMurdie and Holmes used the term ``rarefying''
here and throughout their manuscript. The authors were correct to state
that the distinction between ``rarefying'' and ``rarefaction'' is
confusing and leads to their conflation. Adding to the confusion is that
the papers cited in the first sentence of this quote (i.e., (4--6))
either do not use the words ``rarefy'' or ``rarefying'' or use them
interchangably with ``rarefaction''. For example, Hughes and Hellman did
not use ``rarefy'' and use ``rarefaction'' in the traditional sense with
multiple subsamplings (5). Meanwhile, the QIIME-based literature appears
to use ``rarefy'' and ``rarefaction'' interchangibly meaning only a
single subsampling ((4, 6)). Subsequent researchers have continued to
conflate the results of this study of the effects of rarefying data with
that of traditional rarefaction of data (\textbf{Supplemental Text}). An
exemplar of the confusion is the creation of a technique that uses
``repeatedly rarefying'' as an approach distinct from rarefaction when
they are actually the same thing (10).

As shown in the quoted text from WNWN, McMurdie and Holmes did emphasize
the distinction between rarefying and rarefaction. However, because they
seem to have coined a new definition for rarefying, they added to the
confusion by using the generally used verb form of rarefaction. Further
confusion comes from the author's admonition in the final sentence of
the above quoted text that some researchers have failed to repeat the
subsampling step. Classically, repeating the subsampling step is
rarefaction {[}\textbf{XXXXX}{]}. In other words, subsampling is
rarefaction, but with a single randomization. To minimize confusion, I
will use ``subsamping'' in place of ``rarefying'' thorugh the remainder
of this study.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Select a minimum library size, N\textsubscript{L,m}. Researchers are
  encouraged to report the value of N\textsubscript{L,m}.
\item
  Discard samples that have fewer reads than N\textsubscript{L,m}.
\item
  Subsample the remaining libraries without replacement such that they
  all have size N\textsubscript{L,m}.
\item
  Compute the desired metric (e.g., richness, Shannon diversity,
  Bray-Curtis distances) using the subsampled data
\item
  Repeat steps 3 and 4 a large number of iterations (e.g, 100 or 1,000).
  Researchers are encouraged to report the number of iterations.
\item
  Compute summary statistics (e.g., the mean) using the values from step
  4.
\end{enumerate}

This definition aligns with how rarefaction was originally defined for
comparing richness (i.e., the number of taxa in a community) across
communities when communities are sampled to different depths
{[}\textbf{XXXXX}{]}. With this more general approach to the definition
of rarefaction, rarefaction can be performed using any alpha or beta
diversity metric. This strategy has been widely used by my research
group and others and is available in the mothur software package using
commands such as \texttt{summary.single}, \texttt{rarefaction.single},
\texttt{phylo.diversity}, and \texttt{dist.shared}. The procedure
outlined above could also be used for hypothesis tests of differential
abundance; however, consideration is needed to synthesize the results of
these tests across a large number of replications.

@@

McMurdie and Holmes analyzed the effect of subsampling and other
approaches on clustering accuracy using what they called ``Simulation
A'' in their Figure 2A and elsewhere in their paper. In Simulation A,
they investigated the ability to correctly assign simulated microbiome
samples to one of two clusters representing two simulated treatment
groups. Thankfully, the R code for Simulation A was provided by the
authors in the
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
R-flavored markdown file that was published as Protocol S1 in WNWN. I
will outline the simulation strategy below and will reference line
numbers from their R-flavored markdown file with ``L'' as a prefix.

For the WNWN cluster analysis, sampling distributions for the two
treatment groups were generated using human fecal and ocean data
originally take from the GlobalPaterns dataset (L129) (1). To generate a
fecal and ocean distribution, the authors included any operational
taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3
ocean samples (L60 and L137). The OTUs were sorted by how many of the 7
samples the OTUs were observed in followed by their total abundance
across all 7 samples (L139). From this sorted list they identified the
identifiers of the first 2,000 OTUs (L66). Returning to the 7 samples
they selected the data for the corresponding 2,000 OTUs and pooled the
OTU abundances of the fecal and ocean samples separately to create two
sampling distributions (L144, L159-160, L197-198). Next, the fecal and
ocean distributions were mixed in 8 different fractions to generate two
community types that differend by varying effect sizes (i.e., 1.0, 1.15,
1.25, 1.5, 1.75, 2, 2.5, and 3.5; L170-195, L220); an effect size of 1.0
generated a null model with no difference between the treatment groups.
To simulate the variation in sequencing depth across the 80 samples,
they normalized the number of sequences from each of the 26 samples in
the GlobalPatterns dataset so that the median number of sequences
(Ñ\textsubscript{L}) across the samples had 1,000, 2,000, 5,000, or
10,000 sequences (L324-325). They then randomly sampled the 26
normalized sequencing depths, with replacement, to generate 80 sampling
depths. From each community type, they simulated 40 samples by sampling
to the desired number of sequences (L73, L230-233 and L326-327). Each
simulation condition was repeated 5 times (L85). Finally, they removed
rare and low prevalence OTUs in two steps. First, they removed any OTUs
whose total abundance was less than 3 across all 80 samples and that did
not appear in at least 3 samples (L368-386). Second, they removed any
OTUs that did not have more than 1 sequence in more than 5\% of the 80
samples (i.e., 4 samples) and that did not have a total abundance across
the 80 samples greater than one half of the number of samples in each
community type (i.e., 20) (L523-538, L551). Simulation A consisted of
160 simulations (8 effect sizes x 4 median sampling depths x 5
replicates = 160 simulations).

After generating the OTU counts for the 160 simulations, the authors
applied several normalization methods, distance calculations, and
clustering algorithms to the data. To normalize the OTU data the WNWN
analysis either applied no normalization procedure (L559), calculated
OTU relative abundances (L392-401, L562-563), subsampled the data
(L409-416, L566-569), performed variance stabilization using the DESeq R
package (L478-504, L580-583), or performed Upper-Quartile log-fold
change normalization using the edgeR R package (L425-457, L576-577,
L622-721). In subsampling the data, the authors either included all of
the samples or removed samples whose sequencing depth fell below the 5,
10, 15, 20, 25, and 40 percentile across all 80 samples (L96,
L1065-1077). Non-normalized data were used to calculate Bray-Curtis,
Euclidean, Poisson, and Weighted UniFrac distances (L793-798, L801-806).
Relative abundance data were used to calculate Bray-Curtis, Unweigthed
UniFrac, and Weighted UniFrac distances (L760-767). Subsampled data were
then used as input to calculate distances between samples using
Bray-Curtis, Euclidean, Unweighted UniFrac, and Weighted Unifrac
distances as implemented in the Phyloseq R package, Poisson distances as
implemented in the PoiClaClu R package, and top-mean squared difference
as implemented in the edgeR R package (L769-775, L793-798, L801-806).
DESeq variance stabilization normalized data were used to calculate
Bray-Curtis, Euclidean, and Weighted Unifrac distances (L793-798). The
Upper-Quartile log-fold change normalized data were only used to
calculate top-mean squared difference distances (L801-806). The
resulting distance matrices were used to cluster the 80 samples into one
of two clusters using partitioning around the medoid (PAM), K-means
clustering, and hierarchical clustering (L865-879). Although data for
all three methods were presented in the supplementary Protocol S1, only
the PAM data are presented in the main manuscript. The accuracy of the
clustering assignments was quantified as the fraction of the 80 samples
that were assigned to the correct cluster (L887-908). Since some of the
subsampling conditions removed samples below a minimum sequencing depth
threshold, those samples were counted as mis-clustered samples yielding
minimum accuracies below 50\%.

Although all simulations represent an artificial representation of
reality and can be critiqued, eleven elements of the design of
Simulation A warrant further review.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulated conditions were only replicated 5 times each, potentially
  increasing the sensitivity of results to the choice of the random
  number generator seed
\item
  The average sizes of the libraries were small by modern standards,
  which may limit the generalizability of the results
\item
  DESeq-based variance stabilization was used with distance calculation
  methods that are sensitive to negative values, which likely led to
  nonsense distances and clusters
\item
  A single subsampling of each dataset was evaluated rather than using
  rarefaction, which likely resulted in noisier data
\item
  Results using PAM clustering were not directly compared to those of
  K-means and hierarchical clustering, although close inspection
  suggests that K-means may have been superior to PAM for some
  conditions
\item
  Subsampling removed the smallest 15\% of the samples, which penalized
  accuracy values by 15 percentage points
\item
  The distribution of library sizes was not typical of those commonly
  seen in microbiome analyses, which may limit the generalizability of
  the results
\item
  A filtering step was applied to remove rare taxa from the simulated
  datasets, which may have skewed the shape of the community
  distributions
\item
  There was no accounting for differences in performance when library
  sizes are confounded with treatment group
\item
  Clustering accuracy was used rather than the more direct and
  frequently applied comparisons of beta diversity using permutation
  tests
\item
  There was no consideration of effects of normalization methods on
  alpha diversity metrics, which is the traditional application of
  rarefaction
\end{enumerate}

After replicating the original WNWN simulations, these points will be
evaluated to reassess whether subsampling or rarefaction are
``inadmissible''.

@@

Each simulated condition was replicated 5 times in WNWN and the paper
reports the mean and standard deviation of the replicate clustering
accuracies. The relatively small number of replicates accounts for the
jerkiness of the lines in WNWN's Figures 4 and 5 (e.g.~the Bray-Curtis
distances calculated on the DESeqVS normalized data). A better approach
would have been to use 100 replicates as this would reduce the
dependency of the results on the random number generator's seed. By
increasing the number of replicates it was also possible to compare the
probability of falsely and correctly clustering samples from the same
and different treatment groups together (see points 9-11, below).
Because the accuracies were not symmetric around the mean accuracy
values the median and 95\% confidence intervals or intraquartile range
should have been reported. To test the effect of increasing the number
of replicates, I pulled apart the code in
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
into individual R and bash scripts that were executed using a Snakemake
workflow with the same Conda environment I used above. This was
necessary since the number of simulated conditions increased 20-fold
with the additional replicates. Such intense data processing was not
practical within a single R-flavored markdown document. Again, the
observed results were qualitiatively similar to those generated using
the single R-flavored markdown file (Figures 3 and 4). The increased
number of replications resulted in smoother lines and allowed me to
present empirical 95\% confidence intervals. For all analyses in the
remainder of this paper, I used 100 randomized replicates per condition.

@@

In the 10 years since WNWN was published, sequencing technology has
advanced and sequence collections have grown considerably. For more
modern datasets, it would be reasonable to expect a median number of
sequences larger than 10,000 (see Table 1 of (11)). Therefore, I
included an additional median depth of sampling value of 50,000
sequences with WNWN's four median sequencing sampling depths (i.e.,
1,000, 2,000, 5,000, 10,000). Additional sequencing coverage would be
expected to result in more robust distance values since there would be
more information represented in the data. Indeed, the added sampling
depth showed higher accuracy values at lower effect sizes for the
combinations of normalization methods and distance calculations (Figure
3). Increased sequencing coverage also resulted in improved clustering
accuracy for lower effect sizes when the library size minimum quantile
was decreased (Figure 4). I will revisit the choice of the library size
minimum quantile below.

\hypertarget{deseq-based-variance-stabilization-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values-which-likely-led-to-nonsense-distances-and-clusters}{%
\subsubsection{3. DESeq-based variance stabilization was used with
distance calculation methods that are sensitive to negative values,
which likely led to nonsense distances and
clusters}\label{deseq-based-variance-stabilization-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values-which-likely-led-to-nonsense-distances-and-clusters}}

>>>>

\({\dagger}\) To whom correspondence should be addressed:

@@

In 2014, McMurdie and Holmes published the provocatively titled, ``Waste
not, want not: why rarefying microbiome data is inadmissible''. The
claims of their study have significantly altered how microbiome
researchers control for the unavoidable uneven sequencing depths that
are inherent in modern 16S rRNA gene sequencing. Confusion over the
distinction between the definitions of rarefying and rarefaction
continues to cloud the interpretation of their results. More
importantly, the authors made a variety of choices in designing and
analyzing their simulations that were potentially problematic. I
identified 11 factors that could have compromised the results of the
original study. I reproduced the original simulation results and
assessed the impact of those 11 factors on the underlying conclusion
that rarefying data is inadmissible. Throughout, the design of the
original study made choices that caused rarefying and rarefaction to
appear to perform worse than they truly did. Most important were the
approaches used to assess ecological distances, the removal of samples
with low sequencing depth, and not accounting for conditions where
sampling effort is confounded with treatment group. Although the
original study criticized rarefying for the arbitrary removal of valid
data, repeatedly rarefying data many times (i.e., rarefaction)
incorporates all the data. In contrast, it is the removal of rare taxa
that would appear to remove valid data. Overall, I show that rarefaction
is the most robust approach to control for uneven sampling effort when
considered across a variety of alpha and beta diversity metrics.

\hypertarget{importance}{%
\subsubsection{Importance}\label{importance}}

Over the past 10 years, the best method of normalizing the sequencing
depth of samples characterized by 16S rRNA gene sequencing has been
contentious. An often cited paper by McMurdie and Holmes forcefully
argued that rarefying the number of sequence counts was ``inadmissible''
and should not be employed. However, I identified a number of problems
with the design of their simulations and analysis that could compromise
their results. In fact, when I reproduced and expanded upon their
analysis, it was clear that rarefaction was actually the most robust
approach for controlling for uneven sequencing effort across samples.
Rarefaction limits the rate of falsely detecting and rejecting
differences between treatment groups. Far from being ``inadmissible'',
rarefaction is a valuable tool for analyzing microbiome sequence data.

@@

Microbiome studies that use amplicon sequencing to characterize multiple
samples use PCR to amplify 16S rRNA gene fragments using primers with
distinct barcodes or index sequences (1--3). These barcodes allow
researchers to pool PCR products and then deconvolute the resulting
sequence data based on the barcode sequences. Despite researchers' best
efforts to generate equimolar pools of PCR products, it is common to
observe more than 10-fold variation in the number of sequences per
sample (4). Researchers desire strategies to minimize uneven sequencing
depth and thus need methods to control for this unevenness in their
analyses. Of course, uneven sampling is not unique to microbiome
research and is a challenge faced by all community ecologists (5, 6).
Common approaches to control for the effects of uneven sequencing depths
include use of proportional abundance (i.e., relative abundance),
scaling of counts, parameter estimation, and rarefaction (7--23).

In 2014 Paul McMurdie and Susan Holmes published ``Waste not, want not:
why rarefying microbiome data is inadmissible'' (WNWN) in \emph{PLOS
Computational Biology} (24). Their paper has had a significant impact on
the approaches that microbiome researchers use to analyze 16S rRNA gene
sequence data. According to Google Scholar, this paper has been cited
more than 2,400 times as of May 2023. There has been a rebuttal of WNWN
that showed how rarefaction is beneficial in some cases (25). I have
received correspondence from researchers over the past 10 years asking
how to address critiques from reviewers who criticize my correspondents'
analysis for using rarefaction (e.g., see
\url{https://twitter.com/inanna_nalytica/status/1264299305067786242}). I
have also received similar comments from reviewers regarding my own
work. For example, I received the critique for a preprint that I posted
in 2020 that analyzed the practice of removing rare taxa from microbiome
analyses {[}see comments linked to (4);
\url{https://web.archive.org/web/20230308191505/http://dalmug.org/filter-review/}{]}.
In the process of responding to these reviewers' comments and preparing
a manuscript investigating rarefaction and other approaches to control
for uneven sequencing effort, I grew to appreciate that there is
significant confusion in the field over what is meant by ``rarefying''
and ``rarefaction''.

@@

\begin{quote}
``Instead, microbiome analysis workflows often begin with an ad hoc
library size normalization by random subsampling without replacement, or
so-called rarefying (13, 26, 27). There is confusion in the literature
regarding terminology, and sometimes this normalization approach is
conflated with a non-parametric resampling technique --- called
rarefaction (5), or individual-based taxon re-sampling curves (28) ---
that can be justified for coverage analysis or species richness
estimation in some settings (28), though in other settings it can
perform worse than parametric methods (29). Here we emphasize the
distinction between taxon re-sampling curves and normalization by
strictly adhering to the terms rarefying or rarefied counts when
referring to the normalization procedure, respecting the original
definition for rarefaction. Rarefying is most often defined by the
following steps (13).

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Select a minimum library size, N\textsubscript{L,m}. This has also
  been called the rarefaction level (26), though we will not use the
  term here.
\item
  Discard libraries (microbiome samples) that have fewer reads than
  N\textsubscript{L,m}.
\item
  Subsample the remaining libraries without replacement such that they
  all have size N\textsubscript{L,m}.
\end{enumerate}

@@

It was unfortunate that McMurdie and Holmes used the term ``rarefying''
in this quote and throughout their manuscript. The authors were correct
to state that the distinction between ``rarefying'' and ``rarefaction''
is confusing and leads to their conflation. Adding to the confusion is
that the papers cited in the first sentence of this quote (i.e., (13,
26, 27)) either do not use the words ``rarefy'' or ``rarefying'' or use
them interchangeably with ``rarefaction''. For example, Hughes and
Hellman did not use ``rarefy'' and use ``rarefaction'' in the
traditional sense with multiple subsamplings (13). Meanwhile, the
QIIME-based literature appears to use ``rarefy'' and ``rarefaction''
interchangeably to mean only a single subsampling (26, 27). Subsequent
researchers have continued to conflate the terms when citing WNWN
(\textbf{Supplemental Text}). An exemplar of the confusion is the
creation of a technique that uses ``repeatedly rarefying'' as an
approach distinct from rarefaction when they were re-proposing
traditional rarefaction (30).

As shown in the quoted text from WNWN, McMurdie and Holmes did emphasize
the distinction between rarefying and rarefaction. However, because they
seem to have coined a new definition for rarefying, they added to the
confusion by using the generally used verb form of rarefaction. Further
confusion comes from the author's admonition in the final sentence of
the above quoted text that some researchers have failed to repeat the
subsampling step. Traditionally, repeating the subsampling step a large
number of times and averaging the result is rarefaction (12, 13, 28). In
other words, rarefying or subsampling is rarefaction, but with a single
randomization. To minimize confusion, I will use ``subsampling'' in
place of ``rarefying'' through the remainder of this study.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Select a minimum library size, N\textsubscript{L,m}. Researchers are
  encouraged to report the value of N\textsubscript{L,m}.
\item
  Discard samples that have fewer reads than N\textsubscript{L,m}.
\item
  Subsample the remaining libraries without replacement such that they
  all have size N\textsubscript{L,m}.
\item
  Compute the desired metric (e.g., richness, Shannon diversity,
  Bray-Curtis distances) using the subsampled data
\item
  Repeat steps 3 and 4 a large number of iterations (typically 100 or
  1,000). Researchers are encouraged to report the number of iterations.
\item
  Compute summary statistics (e.g., the mean) using the values from step
  4.
\end{enumerate}

This definition aligns with how rarefaction was originally defined for
comparing richness (i.e., the number of taxa in a community) across
communities when communities are sampled to different depths (5, 6).
With this more general approach to the definition of rarefaction,
rarefaction can be performed using any alpha or beta diversity metric.
This strategy has been widely used by my research group and others and
is available in the mothur software package using commands such as
\texttt{summary.single}, \texttt{rarefaction.single},
\texttt{phylo.diversity}, and \texttt{dist.shared}. The procedure
outlined above could also be used for hypothesis tests of differential
abundance; however, consideration is needed to synthesize the results of
these tests across a large number of replications.

@@

McMurdie and Holmes analyzed the effect of subsampling and other
approaches on clustering accuracy using what they called ``Simulation
A'' in their Figure 2A and elsewhere in WNWN. In Simulation A, they
investigated the ability to correctly assign simulated microbiome
samples to one of two clusters representing two simulated treatment
groups. Thankfully, the R code for Simulation A was provided by the
authors in the
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
R-flavored markdown file that was published as Protocol S1 in WNWN. I
will outline the simulation strategy below and will reference line
numbers from their R-flavored markdown file with ``L'' as a prefix.

For the WNWN cluster analysis, sampling distributions for the two
treatment groups were generated using human fecal and ocean data
originally take from the GlobalPatterns dataset (L129) (2). To generate
the fecal and ocean distributions, the authors included any operational
taxonomic unit (OTU) that appeared in more than one of the 4 fecal and 3
ocean samples (L60 and L137). The OTUs were sorted by how many of the 7
samples the OTUs were observed in followed by their total abundance
across all 7 samples (L139). From this sorted list they obtained the
identifiers of the first 2,000 OTUs (L66). Returning to the 7 samples
they selected the data for the corresponding 2,000 OTUs and pooled the
OTU abundances of the fecal and ocean samples separately to create two
sampling distributions (L144, L159-160, L197-198). Next, the fecal and
ocean distributions were mixed in 8 different fractions to generate two
community types that differed by varying effect sizes (i.e., 1.00, 1.15,
1.25, 1.50, 1.75, 2.00, 2.50, and 3.50; L170-195, L220); an effect size
of 1.00 generated a null model with no difference between the treatment
groups. To simulate the variation in sequencing depth across the 80
samples, they normalized the number of sequences from each of the 26
samples in the GlobalPatterns dataset so that the median number of
sequences (Ñ\textsubscript{L}) across the samples had 1,000, 2,000,
5,000, or 10,000 sequences (L324-325). They then randomly sampled the 26
normalized sequencing depths, with replacement, to generate 80 sampling
depths. From each community type, they simulated 40 samples by sampling
to the desired number of sequences (L73, L230-233 and L326-327). Each
simulation condition was repeated 5 times (L85). Finally, they removed
rare and low prevalence OTUs in two steps. First, they removed any OTUs
whose total abundance was less than 3 across all 80 samples and that did
not appear in at least 3 samples (L368-386). Second, they removed any
OTUs that did not have more than 1 sequence in more than 5\% of the 80
samples (i.e., 4 samples) and that did not have a total abundance across
the 80 samples greater than one half of the number of samples in each
community type (i.e., 20) (L523-538, L551). Simulation A consisted of
160 simulations (8 effect sizes x 4 median sampling depths x 5
replicates = 160 simulations).

After generating the OTU counts for the 160 simulations, the authors
applied several normalization methods, distance calculations, and
clustering algorithms to the data. To normalize the OTU data the WNWN
analysis either applied no normalization procedure (L559), calculated
OTU relative abundances (L392-401, L562-563), subsampled the data
(L409-416, L566-569), performed Variance Stabilization using the
\texttt{DESeq} R package (L478-504, L580-583)(7), or performed
Upper-Quartile Log-Fold Change normalization using the \texttt{edgeR} R
package (L425-457, L576-577, L622-721) (19). In subsampling the data,
the authors either included all of the samples or removed samples whose
sequencing depth fell below the 5, 10, 15, 20, 25, and 40 percentile
across all 80 samples (L96, L1065-1077).

Non-normalized data were used to calculate Bray-Curtis, Euclidean,
Poisson, and Weighted UniFrac distances (L793-798, L801-806). Relative
abundance data were used to calculate Bray-Curtis, Unweighted UniFrac ,
and Weighted UniFrac distances (L760-767). Subsampled data were used as
input to calculate distances between samples using Bray-Curtis,
Euclidean, Unweighted UniFrac, Weighted UniFrac distances, Poisson
distances, and top-Mean Squared Difference (L769-775, L793-798,
L801-806). DESeq Variance Stabilization normalized data were used to
calculate Bray-Curtis, Euclidean, and Weighted Unifrac distances
(L793-798). The Upper-Quartile Log-Fold Change normalized data were only
used to calculate top-mean squared difference distances (L801-806). WNWN
calculated Bray-Curtis (31), Euclidean (31), Unweighted UniFrac (32),
and Weighted UniFrac distances (33) using the \texttt{Phyloseq} R
package (34). They calculated Poisson distances using the
\texttt{PoiClaClu} R package (35). The top-mean squared difference was
calculated using the \texttt{edgeR} R package (19). The resulting
distance matrices were used to cluster the 80 samples into one of two
clusters using partitioning around the medoid (PAM), K-means clustering,
and hierarchical clustering (L865-879). Although data for all three
methods were presented in Protocol S1, only the PAM data were presented
in the main manuscript. The accuracy of the clustering assignments was
quantified as the fraction of the 80 samples that were assigned to the
correct cluster (L887-908). Since some of the subsampling conditions
removed samples below a minimum sequencing depth threshold, those
samples were counted as mis-clustered samples yielding minimum
accuracies below 50\%.

Although all simulations represent an artificial representation of
reality and can be critiqued, I have identified eleven elements of the
design of Simulation A that warranted further review.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Simulated conditions were only replicated 5 times each, potentially
  increasing the sensitivity of results to the choice of the random
  number generator seed
\item
  The average sizes of the libraries were small by modern standards,
  which may limit the generalizability of the results
\item
  DESeq-based Variance Stabilization generates negative values and was
  used with distance calculation methods that are sensitive to negative
  values, which likely led to nonsense distances and clusters
\item
  A single subsampling of each dataset was evaluated rather than using
  rarefaction, which likely resulted in noisier data
\item
  Results using PAM clustering were not directly compared to those of
  K-means and hierarchical clustering, although close inspection
  suggests that K-means may have been superior to PAM for some
  conditions
\item
  Subsampling removed the smallest 15\% of the samples, which penalized
  accuracy values by 15 percentage points
\item
  The distribution of library sizes was not typical of those commonly
  seen in microbiome analyses, which may limit the generalizability of
  the results
\item
  A filtering step was applied to remove rare taxa from the simulated
  datasets, which may have skewed the shape of the community
  distributions
\item
  There was no accounting for differences in performance when library
  sizes are confounded with treatment group
\item
  Clustering accuracy was used rather than the more direct and
  frequently applied comparisons of beta diversity using permutation
  tests
\item
  There was no consideration of effects of normalization methods on
  alpha diversity metrics, which is the traditional application of
  rarefaction
\end{enumerate}

Below, I replicated the original WNWN simulations and evaluated these
points to reassess whether subsampling or rarefaction are
``inadmissible''.

@@

Each simulated condition was replicated 5 times in WNWN and the paper
reported the mean and standard deviation of the replicate clustering
accuracies. The relatively small number of replicates accounts for the
jerkiness of the lines in WNWN's Figures 4 and 5 (e.g.~the Bray-Curtis
distances calculated on the DESeqVS normalized data). A better approach
would have been to use 100 replicates as this would reduce the
dependency of the results on the random number generator's seed. By
increasing the number of replicates it was also possible to compare the
probability of falsely and correctly clustering samples from the same
and different treatment groups together (see points 10 and 11, below).
Because the accuracies were not symmetric around the mean accuracy
values the median and 95\% confidence intervals or interquartile range
should have been reported. To test the effect of increasing the number
of replicates, I pulled apart the code in
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
into individual R and bash scripts that were executed using a Snakemake
workflow with the same Conda environment I used above. This was
necessary since the number of simulated conditions increased 20-fold
with the additional replicates. Such intense data processing was not
practical within a single R-flavored markdown document. Again, the
observed results were qualitatively similar to those generated using the
single R-flavored markdown file (Figures 3 and 4). The increased number
of replications resulted in smoother lines and allowed me to present
empirical 95\% confidence intervals. For all analyses in the remainder
of this paper, I used 100 randomized replicates per condition.

@@

In the 10 years since WNWN was published, sequencing technology has
advanced and sequence collections have grown considerably. For more
modern datasets, it would be reasonable to expect a median number of
sequences larger than 10,000 (see Table 1 of (4)). Therefore, I included
an additional median depth of sampling value of 50,000 sequences with
WNWN's four median sequencing sampling depths (i.e., 1,000, 2,000,
5,000, 10,000). Additional sequencing coverage would be expected to
result in more robust distance values since there would be more
information represented in the data. Indeed, the added sampling depth
showed higher accuracy values at lower effect sizes for the combinations
of normalization methods and distance calculations (Figure 3). Increased
sequencing coverage also resulted in improved clustering accuracy for
lower effect sizes when the library size minimum quantile was decreased
(Figure 4). I will revisit the choice of the library size minimum
quantile below.

\hypertarget{deseq-based-variance-stabilization-generates-negative-values-and-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values-which-likely-led-to-nonsense-distances-and-clusters}{%
\subsubsection{3. DESeq-based Variance Stabilization generates negative
values and was used with distance calculation methods that are sensitive
to negative values, which likely led to nonsense distances and
clusters}\label{deseq-based-variance-stabilization-generates-negative-values-and-was-used-with-distance-calculation-methods-that-are-sensitive-to-negative-values-which-likely-led-to-nonsense-distances-and-clusters}}

Close comparison of WNWN's Figure 4 and my version (Figure 3) revealed
several difference between the two plots. First, in the WNWN analysis,
the accuracies for the Weighted UniFrac distances at the largest effect
size (i.e., 3.5) were 1.00 for median sequencing depths of 1,000, 2,000,
and 10,000. In my version of the analysis, the values for the same
sequencing depths were 1.00, 1.00, and 1.00, respectively. The 95\%
confidence interval for these accuracies spanned between 0.51 and 1.00.
Second, clustering accuracy for Bray-Curtis and Weighted UniFrac
distances calculated using DESeq Variance Stabilization normalized OTU
counts were different between the original and my simulation at smaller
effect sizes and had wide confidence intervals. Inspection of the DESeq
Variance Stabilization normalized OTU counts revealed that the method
resulted in negative values. It has been suggested that WNWW turned
negative DESeq normalized counts to zero (25); however, I was unable to
find code in \texttt{simulation-cluster-accuracy-server.Rmd} that made
this transformation. In fact, rendering the R-flavored markdown files in
WNWN's Protocol S1 generated warning messages when passing the DESeq
normalized counts to the Bray-Curtis calculator, which said, ``results
may be meaningless because data have negative entries in method
`bray'\,''. Although the Weighted UniFrac calculator function did not
generate a similar warning message, negative count values would also
result in similarly meaningless distances. Both are due to the fact that
the distance calculators sum the counts of each OTU in both samples
being compared. In contrast, a Euclidean distance does not use a similar
sum, but sums the square of the difference between the OTU abundance in
each sample. Even if negative values were converted to zeroes, this
would effectively the same as removing rare taxa, which could have a
significant impact on the shape of the communities (4). To assess the
prevalence of negative counts in the simulated data, I quantified the
fraction of negative values in the OTU matrix from each simulation and
counted the number of simulations where the normalized OTU table had at
least one negative value (Figure 5). In general the fraction of negative
OTU counts increased with effect size, but decreased with sequencing
effort. The fraction of simulations with at least one negative value
increased with effect size and sequencing effort. The high frequency of
negative OTU counts resulted in highly variable Bray-Curtis and Weighted
UniFrac values. It is likely that because the WNWN analysis only used 5
replicates that the large variation in accuracies at high effect sizes
was missed initially. For the rest of this reanalysis study, I will only
report results using the DESeq-based Variance Stabilization
normalization with the Euclidean distance.
